Inefficient Concessions and Mediation

### 2021-01-29 draft.tex
Fill out mediation theorem, but see a mistake in high type IC (to be fixed next)

%Concessions form the centerpiece of the GRIT (graduated and reciprocated initiatives in tension reduction) theory in the psychological literature (Osgood 1962) and have shown ample success in lab settings. Komorita (1973) conducts lab experiments using a repeated Prisoner's Dilemma similar to the setup in our paper and finds costly conciliatory acts (concessions) to be useful in promoting cooperation. De Dreu (1995) likewise uses a lab setting to experiment with concessions and conflict.  %de dreu bases experiments on shelling 60 and bacharach and lawler 81/ lawler 92 about deterrance, the shelling conflict spiral was not fully rejected, but the alternative of deterrance was supported
%In addition to concessions, De Dreu also reports laboratory results on mediation; he concludes that a weaker mediator should work with promises while a more powerful one can be effective with threats.  Larson (1988) and others have applied this framework to political science questions. On the theoretical side, Filson and Werner (2007) use a bargaining setup to explore the sensitivities of costs compared to the sensitivity of giving concessions. 
%LAKE: The psychological literature as applied in political science spent a lot of time in the 1970s and 1980s focusing on Graduated Reductions in Tensions (GRIT) as a form of confidence building. See Debbie LarsonÕs work (sheÕs at Poli Sci, UCLA, I donÕt have the cites handy). This is clearly relevant to your discussion of the laboratory experiments on p.3. BH WATSON'S STARTING SMALL COULD BE RELEVANT HERE

@@

Beginning in period 1, countries engage in an infinitely repeated Prisoner's Dilemma once per period with stage game payoffs in Table 1 below. While not a good representation of all-out war or even a zero-sum short term war, the Prisoner's Dilemma is appropriate to represent a limited, continued conflict. In cases appropriately modeled by a Prisoner's Dilemma, it would be better for both parties to exit the protracted conflict, but no party has the unilateral incentive to do so. The critical results of this paper do not depend on a prisoner's dilemma structure; they require only that a better result can be achieved through trust or credible commitment. Thus, a stag hunt or other game structures are also possible. The Prisoner's Dilemma is used because of its wide exposure in the literature.\footnote{This payoff matrix does not distinguish between civil and interstate conflict, though in practice the effects on the payoffs might indeed have a different structure in the different cases.} %cite difference from fearon here?
%LAKE More generally, the problem of war, on which you implicitly focus, is now understood as a pure bargaining game (i.e., zero sum). This is not well represented by the PD. You need to make the link Ð or differentiate Ð what youÕre doing here from the Fearon-Powell models to which you appeal.

@@

\begin{lemma}
	If both the \emph{concessions separating equilibrium with efficient concessions} and the \emph{concessions separating equilibrium with inefficient concessions} exist, the \emph{concessions separating equilibrium with efficient gifts} is preferred when both the probability of encountering a high type and the benefit from making peace are sufficiently large.
	\label{lemma:efficiency}
\end{lemma}
\emph{Proof:} See Appendix.

%need to also write constraints for high type IC to be satisfied, or appeal to forward induction
%? And, a mechanism improves upon inefficiency because it allows a high types to only give a gift to other high types, thus decreasing their costs and still giving the same equilibrium matches.  . Does this rely on varied costs of giving- i.e with same costs can a mechanism still improve? 
%note damages due to alpha can be understood as political losses/ embarrassment of failure to make peace as much as they can be seen as military damages.
%{\bf Theorem 12:}\\  
%When concessions are inefficient, $M$ can still improve over bilateral negotiation. \\
% \emph{Proof}: A mechanism works as in proof (6) with the additional liability that high-high concession exchanges are no longer efficient so a mechanism's advantage is not as strong.

The intuition for Lemma~\ref{lemma:efficiency} is that future material value of concession helps you when your negotiating partner turns out to be a high type, and this happens $p$ of the time. On other other hand, the material value of your concession \emph{hurts} you when your negotiating partner turns out to be a low type. Similarly, the larger are the benefits from Trust, the larger the welfare gain when the negotiating partner is a high type and the larger are the damages from Fight, the larger is the welfare loss when the negotiating partner turns out to be a low type.

@@

\begin{theorem}
\emph{Mediation}
\begin{enumerate}
		\item[(a)] The mediation mechanism $M$ is incentive compatible when the specified concession is at least as large as $\frac{\left(1 - \de_l \right)\left(T +D \right)}{p - e\left(p +(1-\de_l)T -W \right)}$.
		\item[(b)] In the best concessions separating equilibrium when gifts have full material value, high types give a concession of $\frac{\left(1-\de_l\right)p\left(D + T\right)}{\left(1-\de_l\right)\left(1 - pT \right) + (1-p) D}$, which is the smallest concession necessary to separate. Equilibria with higher concessions yield strictly lower payoffs.
\end{enumerate}
	\label{theorem:med}
\end{theorem}
\emph{Proof:} See Appendix.

high type IC:
$$pX_{TT}^h+(1-p)X_{FF}^h - pg_h +peg_h \geq X_{FF}^h $$

@@

This paper shows that inefficient concessions may be used to achieve peace when efficient concessions are not sufficient to do so; it also demonstrates how a mediator can achieve peace in these situations without resorting to inefficient concessions. Opportunities for further research clearly include looking at the roles of different types of mediation. Because of the many humanitarian, security and diplomatic implications of this work, both deeper and broader analysis is required.
%This said, if the paper is really about mediation, then the mediator needs more scrutiny. What are the mediatorÕs incentives? Why should he reveal truthfully what the parties tell him? It would seem simple enough to make the mediator part of the game, with an payment function that is a fraction of the surplus from eliminating the inefficiency. This leads to interesting implications. Given some cost to the mediator, they will be more likely to volunteer when e* is large, etc. But treating the mediator as a deus ex machina leaves the current model incomplete.

@@

Here was have assumed that the mediator specifies $e=1$, which means that gift giving and receiving cancels out. The left hand side has zero gifts because the low-type player under consideration has truthfully revealed itself to be a low type. The mediator only specifies strictly positive gifts if there are two reports of `High.' Likewise, there are no gifts in the $(1-p)X_{FF}^l$ term on the right because the negotiating partner has made a report of `Low.' This leaves
		$$X_{FF}^l \geq X_{FT}^l.$$
		Expanding, we have
		\[
			\frac{W-D}{1-\de_l} \geq T + \left(\cdot \right) + W + \left(\cdot \right) + \frac{\de_l}{1-\de_l}\left(W+ \left(\cdot \right) -D - \left(\cdot \right) \right)
		\]
There are no gifts on the left hand side and so the payoffs from Table 1 apply. The right hand side has place holders $\left(\cdot\right)$ for the impact of the material value of gifts. On the right hand side, we have only the case where both parties declare themselves to be high types. Both parties are instructed to give efficient gifts. The negotiating partner really \emph{is} a high type and thus invests entirely in civil society because they expect to play the ``Peace'' equilibrium. The low-type player under consideration knows it will defect and end up in a ``War'' equilibrium and so invests entirely in the military. 
%		\begin{multline}
%			W-D \geq 
%			\left(1-\de_l\right) \left[T + \al_2Tg^M + W + \left(1-\al_1\right)Wg_2 \right] \\
%			+ \de_l \left(W + \left(1-\al_1\right)Wg_2-D - \left(1-\al_2\right)Dg^M\right)
%			\label{eq:lowICex}
%		\end{multline}
		\begin{equation}
			W-D \geq 
			\left(1-\de_l\right) \left[T + Tg_h + W + Wg_h \right] + \de_l \left(W + Wg_h -D \right)
			\label{eq:med}
		\end{equation}
Canceling like terms and rearranging, we have
			\[
					0 \geq \left(1 - \de_l \right) T + \left(1 - \de_l \right) Tg_h + Wg_h + \left(1 - \de_l \right) D
			\]
			Since we assume throughout that gifts must be non-negative, each of $T, W$ and $D$ are positive and $0 \leq \de_l \leq 1$, this inequality can never hold.	\hfill $\blacksquare$\\
			\\
\emph{Proof of Theorem~\ref{theorem:med}:}\\
For the general form of the low type incentive compatibility constraint, we start from Expression~\ref{eq:med} and add back in the $e$'s that were removed when assuming $e=1$ in the proof of Theorem~\ref{theorem:med}. So we start from
\begin{equation*}
			W-D \geq 
			\left(1-\de_l\right) \left[T + Teg_h + W + Weg_h \right] + \de_l \left(W + Weg_h -D \right) + peg_h - pg_h
		\end{equation*}
Simplifying, we have
\[
	pg_h\left(1 - e\right) \geq \left(T - \de_l T \right) \left(1 + eg_h\right) + Weg_h + \left(1 - \de_l \right) D
\]
Gathering the terms involving $g_h$ on the left hand side:
\[
	g_h\left[p\left(1 - e\right) -(1-\de_l)eT -eW \right] \geq \left(T - \de_l T \right) + \left(1 - \de_l \right) D
\]
Isolating $g_h$ on the left hand side and gathering terms, we have 
\begin{equation}
	g_h \geq \frac{\left(1 - \de_l \right)\left(T +D \right)}{p - e\left(p +(1-\de_l)T -W \right)} = g^M
	\label{ex:med3}
\end{equation}
As before, the optimal gift is the one where Expression~\ref{ex:med3} holds with equality. We call this value $g^M$. We see here that $g^M$ depends on $e$, which is also the choice of the mediator \emph{and} that we must have $\overline{e} = p - e \left( p + T - \de_l T + W \right) > 0$ in order for the optimal concession to be both positive and finite. Both the optimal concession increases toward infinity as $e$ approaches $\frac{p}{p + \left(1-\de_l\right) + W}$, which is the value that makes the denominator of $g^M$ equal to 0.

>>>>

%Concessions form the centerpiece of the GRIT (graduated and reciprocated initiatives in tension reduction) theory in the psychological literature (Osgood 1962) and have shown ample success in lab settings. Komorita (1973) conducts lab experiments using a repeated Prisoner's Dilemma similar to the setup in our paper and finds costly conciliatory acts (concessions) to be useful in promoting cooperation. De Dreu (1995) likewise uses a lab setting to experiment with concessions and conflict.  %de dreu bases experiments on shelling 60 and bacharach and lawler 81/ lawler 92 about deterrance, the shelling conflict spiral was not fully rejected, but the alternative of deterrance was supported
%In addition to concessions, De Dreu also reports laboratory results on mediation; he concludes that a weaker mediator should work with promises while a more powerful one can be effective with threats.  Larson (1988) and others have applied this framework to political science questions. On the theoretical side, Filson and Werner (2007) use a bargaining setup to explore the sensitivities of costs compared to the sensitivity of giving concessions. 
%LAKE: The psychological literature as applied in political science spent a lot of time in the 1970s and 1980s focusing on Graduated Reductions in Tensions (GRIT) as a form of confidence building. See Debbie Larson’s work (she’s at Poli Sci, UCLA, I don’t have the cites handy). This is clearly relevant to your discussion of the laboratory experiments on p.3. BH WATSON'S STARTING SMALL COULD BE RELEVANT HERE

@@

Beginning in period 1, countries engage in an infinitely repeated Prisoner's Dilemma once per period with stage game payoffs in Table 1 below. While not a good representation of all-out war or even a zero-sum short term war, the Prisoner's Dilemma is appropriate to represent a limited, continued conflict. In cases appropriately modeled by a Prisoner's Dilemma, it would be better for both parties to exit the protracted conflict, but no party has the unilateral incentive to do so. The critical results of this paper do not depend on a prisoner's dilemma structure; they require only that a better result can be achieved through trust or credible commitment. Thus, a stag hunt or other game structures are also possible. The Prisoner's Dilemma is used because of its wide exposure in the literature.\footnote{This payoff matrix does not distinguish between civil and interstate conflict, though in practice the effects on the payoffs might indeed have a different structure in the different cases.} %cite difference from fearon here?
%LAKE More generally, the problem of war, on which you implicitly focus, is now understood as a pure bargaining game (i.e., zero sum). This is not well represented by the PD. You need to make the link – or differentiate – what you’re doing here from the Fearon-Powell models to which you appeal.

@@

\begin{lemma}
	If the \emph{concessions separating equilibrium with efficient concessions} exists, it is preferred to the \emph{concessions separating equilibrium with inefficient gifts} when both the probability of encountering a high type and the benefit from making peace are sufficiently large.
	\label{lemma:efficiency}
\end{lemma}
\emph{Proof:} See Appendix.

%need to also write constraints for high type IC to be satisfied, or appeal to forward induction
%? And, a mechanism improves upon inefficiency because it allows a high types to only give a gift to other high types, thus decreasing their costs and still giving the same equilibrium matches.  . Does this rely on varied costs of giving- i.e with same costs can a mechanism still improve? 
%note damages due to alpha can be understood as political losses/ embarrassment of failure to make peace as much as they can be seen as military damages.
%{\bf Theorem 12:}\\  
%When concessions are inefficient, $M$ can still improve over bilateral negotiation. \\
% \emph{Proof}: A mechanism works as in proof (6) with the additional liability that high-high concession exchanges are no longer efficient so a mechanism's advantage is not as strong. 
The intuition for Lemma~\ref{lemma:efficiency} is that future material value of concession helps you when your negotiating partner turns out to be a high type, and this happens $p$ of the time. On other other hand, the material value of your concession \emph{hurts} you when your negotiating partner turns out to be a low type. Similarly, the larger are the benefits from Trust, the larger the welfare gain when the negotiating partner is a high type and the larger are the damages from Fight, the larger is the welfare loss when the negotiating partner turns out to be a low type.

@@

\begin{theorem}
\emph{Mediation}
\begin{enumerate}
		\item[(a)] The mediation mechanism $M$ is incentive compatible for the low type when the specified concession is greater than or equal to $\frac{\left(1 - \de_l \right)\left(T +D \right)}{p - e\left(p +(1-\de_l)T +W \right)}$ and $p > e\left(p+\left(1-\de_l\right)T + W\right)$.
		\item[(b)] The mediation mechanism $M$ is incentive compatible for the high type when
			\begin{itemize}
				\item[(i)] $\de_h(T+D)\geq W$ and $e(T-W) >1 -\de_h$;
				\item[(ii)] $\de_h(T+D)\geq W$, $e(T-W) <1 -\de_h$ and the specified concession is less than or equal to $\frac{\de_h(T+D)-W}{1-\de_h - e\left(T +W \right)}$; or
				\item [(iii)] $\de_h(T+D)\leq W$, $e(T-W) >1 -\de_h$ and the specified concession is greater than or equal to $\frac{\de_h(T+D)-W}{1-\de_h - e\left(T +W \right)}$.
			\end{itemize}
		\item[(c)] The mediation mechanism $M$ is always individually rational for the low type. If $Te > (1-\de_h)(1-e)$, the high type's individual rationality constraint implies an additional restriction that the specified concession is less than or equal to $\frac{ T + D - W}{Te - (1-\de_h)(1-e)}$.
\end{enumerate}
	\label{theorem:med}
\end{theorem}
\emph{Proof:} See Appendix.

@@

This paper shows that inefficient concessions may be used to achieve peace when efficient concessions are not sufficient to do so; it also demonstrates how a mediator can achieve peace in these situations without resorting to inefficient concessions. Opportunities for further research clearly include looking at the roles of different types of mediation. Because of the many humanitarian, security and diplomatic implications of this work, both deeper and broader analysis is required.
%This said, if the paper is really about mediation, then the mediator needs more scrutiny. What are the mediator’s incentives? Why should he reveal truthfully what the parties tell him? It would seem simple enough to make the mediator part of the game, with an payment function that is a fraction of the surplus from eliminating the inefficiency. This leads to interesting implications. Given some cost to the mediator, they will be more likely to volunteer when e* is large, etc. But treating the mediator as a deus ex machina leaves the current model incomplete.

@@

Here was have assumed that the mediator specifies $e=1$, which means that gift giving and receiving cancels out. The left hand side has zero gifts because the low-type player under consideration has truthfully revealed itself to be a low type. The mediator only specifies strictly positive gifts if there are two reports of `High.' Likewise, there are no gifts in the $(1-p)X_{FF}^l$ term on the right because the negotiating partner has made a report of `Low.' This leaves
		$$X_{FF}^l \geq X_{FT}^l.$$
		Expanding, we have
		\[
			\frac{W-D}{1-\de_l} \geq T + \left(\cdot \right) + W + \left(\cdot \right) + \frac{\de_l}{1-\de_l}\left(W+ \left(\cdot \right) -D - \left(\cdot \right) \right)
		\]
There are no gifts on the left hand side and so the payoffs from Table 1 apply. The right hand side has place holders $\left(\cdot\right)$ for the impact of the material value of gifts. On the right hand side, we have only the case where both parties declare themselves to be high types. Both parties are instructed to give efficient gifts. The negotiating partner really \emph{is} a high type and thus invests entirely in civil society because they expect to play the ``Peace'' equilibrium. The low-type player under consideration knows it will defect and end up in a ``War'' equilibrium and so invests entirely in the military. 
%		\begin{multline}
%			W-D \geq 
%			\left(1-\de_l\right) \left[T + \al_2Tg^M + W + \left(1-\al_1\right)Wg_2 \right] \\
%			+ \de_l \left(W + \left(1-\al_1\right)Wg_2-D - \left(1-\al_2\right)Dg^M\right)
%			\label{eq:lowICex}
%		\end{multline}
			W-D \geq 
			\left(1-\de_l\right) \left[T + Tg_h + W + Wg_h \right] + \de_l \left(W + Wg_h -D \right)
			\label{eq:med}
Canceling like terms and rearranging, we have
			\[
					0 \geq \left(1 - \de_l \right) T + \left(1 - \de_l \right) Tg_h + Wg_h + \left(1 - \de_l \right) D
			\]
			Since we assume throughout that gifts must be non-negative, each of $T, W$ and $D$ are positive and $0 \leq \de_l \leq 1$, this inequality can never hold.	\hfill $\blacksquare$\\
\emph{Proof of Theorem~\ref{theorem:med}:}\\
(a) For the general form of the low type incentive compatibility constraint, we start from Expression~\ref{eq:med} and add back in the $e$'s that were removed when assuming $e=1$ in the proof of Theorem~\ref{theorem:med}. So we start from
			W-D \geq 
			\left(1-\de_l\right) \left[T + Teg_h + W + Weg_h \right] + \de_l \left(W + Weg_h -D \right) + peg_h - pg_h
\[
	pg_h\left(1 - e\right) \geq \left(T - \de_l T \right) \left(1 + eg_h\right) + Weg_h + \left(1 - \de_l \right) D
\]
Gathering the terms involving $g_h$ on the left hand side:
\[
	g_h\left[p\left(1 - e\right) -(1-\de_l)eT -eW \right] \geq \left(T - \de_l T \right) + \left(1 - \de_l \right) D
\]
The right hand side is positive. If the left hand side were negative, we would have a negative upper bound on the size of the gift. Given the requirement that the gift be non-negative, the low type's incentive constraint cannot be satisfied  in this case. Therefore we must assume that $p > e\left(p+\left(1-\de_l\right)T + W\right)$. Using this to isolate $g_h$ on the left hand side and gathering terms, we have 
	g_h \geq \frac{\left(1 - \de_l \right)\left(T +D \right)}{p - e\left(p +(1-\de_l)T -W \right)}
	\label{eq:med3}
If Condition~\ref{eq:med3} is met, the mechanism is incentive compatible for the low type.\\
(b) We now turn to the high type, whose incentive compatibility constraint is
$$pX_{TT}^h+(1-p)X_{FF}^h - pg_h +peg_h \geq X_{FF}^h $$
It is always optimal for the high type to invest in civil society when they truthfully reveal and to invest in the military when they lie about their type. Thus the constraint becomes
  \frac{p}{1-\de_h}\left(T + Teg \right) - pg +peg+ \frac{1-p}{1-\de_h}\left(W-D\right) \geq \\
	p\left(T + W + Weg +eg + \frac{\de_h}{1-\de_h}\left(W + Weg -D\right)\right) + \frac{1-p}{1-\de_h}\left(W-D\right)
This simplifies to 
  \frac{1}{1-\de_h}\left(T + Teg \right) - g \geq T + W + Weg + \frac{\de_h}{1-\de_h}\left(W + Weg -D\right)
%  T + Teg - ( 1-\de_h)g \geq ( 1-\de_h)\left(T + W + Weg\right) + \de_h\left(W + Weg -D\right)
  T + Teg - ( 1-\de_h)g \geq ( 1-\de_h)T + W + Weg - \de_h D
%  Teg - Weg -( 1-\de_h)g \geq -\de_h T + W - \de_h D
  g\left(Te - We -( 1-\de_h)\right) \geq W - \de_h (D+T)
	\label{eq:med4}
We have four cases to consider depending on signs:
\begin{itemize}
	\item [(i)] When $\de_h(T+D)\geq W$ and $e(T-W) >1-\de_h$, then we have a lower bound on the size of the gift that is non-positive, i.e. no additional restriction beyond the restrictions needed to satisfy the low type incentive compatibility constraint.
	\item [(ii)] When $\de_h(T+D)\geq W$ and $e(T-W) <1-\de_h$, Expression~\ref{eq:med4} is a non-negative upper bound on the size of the gift.
	\item [(iii)] When $\de_h(T+D)\leq W$ and $e(T-W) >1-\de_h$, Expression~\ref{eq:med4} is a non-negative lower bound on the size of the gift, which may be larger or smaller than the bound in Expression~\ref{eq:med3}. This is an additional restriction beyond that for the low type only when the bound is larger than the one in Expression~\ref{eq:med3}.
	\item [(iv)] When $\de_h(T+D)\leq W$ and $e(T-W) <1-\de_h$, Expression~\ref{eq:med4} is a non-positive upper bound on the size of the gift and therefore there is no incentive compatible mechanism.
\end{itemize}
(c): The low type's individual rationality constraint is satisfied trivially: the low type gets the `War' payoff forever outside the mechanism; it also gets the `War' payoff forever \emph{inside} the mechanism while not giving or receiving any concession.

For the high type, we need
\begin{equation*}
  \frac{p}{1-\de_h}\left(T + Teg \right) - pg +peg+ \frac{1-p}{1-\de_h}\left(W-D\right) \geq \frac{1}{1-\de_h}\left(W-D\right)
\end{equation*}
This simplifies to
\begin{equation*}
  \frac{1}{1-\de_h}\left(T + Teg \right) - g +eg \geq \frac{1	}{1-\de_h}\left(W-D\right)
\end{equation*}
\begin{equation*}
  T + Teg - (1-\de_h)(1-e)g \geq W-D
\end{equation*}
\begin{equation*}
  T + Teg - (1-\de_h)(1-e)g \geq W-D
\end{equation*}
\begin{equation*}
  T + D - W \geq \left[Te - (1-\de_h)(1-e)\right]g 
\end{equation*}
If $Te < (1-\de_h)(1-e)$, this is a negative lower bound on the size of the concession and thus implies no additional restriction. If $Te = (1-\de_h)(1-e)$, there is again no restriction because $T+D >W$ by assumption. If $Te > (1-\de_h)(1-e)$, the high type's individual rationality constraint requires that $\frac{ T + D - W}{Te - (1-\de_h)(1-e)}>g$.

\hfill $\blacksquare$\\

As before, the optimal gift is the one where Expression~\ref{eq:med3} holds with equality [NOT NECESSARILY BECAUSE OF HIGH TYPE IC]. We call this value $g^M$. We see here that $g^M$ depends on $e$, which is also the choice of the mediator \emph{and} that we must have $\overline{e} = p - e \left( p + T - \de_l T + W \right) > 0$ in order for the optimal concession to be both positive and finite. Both the optimal concession increases toward infinity as $e$ approaches $\frac{p}{p + \left(1-\de_l\right) + W}$, which is the value that makes the denominator of $g^M$ equal to 0.
